\documentclass[12pt]{article}
\usepackage[a4paper,margin=2.5cm]{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    urlcolor=blue,
    citecolor=blue,
    pdfinfo={DOI={10.6084/m9.figshare.31122970}}
}

\setstretch{1.2}

\title{Validity-Aware AI: An Entropy-Bounded Architecture for Regime-Sensitive Inference\\[1em]
\large Unified Architecture Statement\\[0.5em]
\normalsize Entropy, Validity, and Regime-Aware Systems}

\author{Morten Magnusson\\
\small Independent Researcher\\
\small ORCID: 0009-0002-4860-5095}

\date{January 2026\\[0.5em]
\small DOI: 10.6084/m9.figshare.31122970}

\begin{document}

\maketitle

\begin{abstract}
This document provides a canonical reference for the integration of Energy-Flow Cosmology (EFC), Entropy-Bounded Empiricism (EBE), the L0--L3 regime architecture, and Validity-Aware AI. The unified architecture spans four distinct but non-overlapping layers: ontology, epistemology, regime dynamics, and operational implementation. Across all domains, the architecture enforces a single invariant: no system may generate confident claims outside the regime where such claims are empirically valid.
\end{abstract}

\noindent\textbf{Keywords:} unified architecture, entropy-bounded empiricism, regime architecture, validity awareness, EFC, EBE, L0--L3

\vspace{1em}

This work presents a unified architecture for understanding \textbf{validity, inference, and system behavior} across physical, informational, and artificial domains. The architecture integrates four distinct but non-overlapping layers: ontology, epistemology, regime dynamics, and operational implementation.

\section{Ontological Layer (S0--S1)}

At the most fundamental level, the framework assumes an ontological distinction between:

\begin{itemize}
    \item \textbf{S0}: the total space of possible structures and dynamics
    \item \textbf{S1}: the realized universe or system instantiated within that space
\end{itemize}

This layer defines \textit{what can exist} and \textit{what does exist}. It does not prescribe dynamics, inference, or models. It serves solely as the container within which all other layers operate.

\section{Epistemological Layer (Entropy-Bounded Empiricism, EBE)}

Entropy-Bounded Empiricism defines the conditions under which inference is valid within S1. Its core principle is:

\begin{quote}
\textit{Inference is valid only within regimes where data and dynamics remain consistent.}
\end{quote}

EBE treats \textbf{entropy as a limiting factor on empirical validity}, not as a property of models but of the underlying system structure. It specifies where empirical claims can be made, where uncertainty must be made explicit, and where inference must cease.

EBE is domain-agnostic: it applies equally to cosmology, biology, cognition, economics, and artificial systems.

\section{Regime Architecture Layer (L0--L3)}

Within the epistemic boundaries defined by EBE, system behavior is structured by a universal regime architecture:

\begin{itemize}
    \item \textbf{L0 (Latent)}: potential structure without active dynamics
    \item \textbf{L1 (Stable Active)}: low-entropy regimes where predictive inference is valid
    \item \textbf{L2 (Complex Active)}: high-entropy regimes with contested dynamics and non-linear behavior
    \item \textbf{L3 (Residual)}: preserved structure without active, predictive dynamics
\end{itemize}

Regime transitions are \textbf{phase-like}, extended in time, and driven by entropy crossing domain-specific thresholds. L3 is not disorder, but residual structure: memory without process.

L0--L3 define \textit{how systems evolve and degrade}, not what they are.

\section{Physical Implementation (Energy-Flow Cosmology, EFC)}

Energy-Flow Cosmology provides a concrete physical realization of this architecture in the cosmological domain. In EFC:

\begin{itemize}
    \item entropy gradients govern structure formation
    \item regime transitions explain where standard models succeed or fail
    \item empirical validation is explicitly regime-dependent
\end{itemize}

EFC is not the foundation of the architecture, but its \textbf{physical instantiation} in our universe.

\section{Informational and AI Implementation (Validity-Aware AI)}

In artificial systems, the same architecture manifests as \textbf{regime-aware inference} over structured knowledge.

\begin{itemize}
    \item Entropy is measured in knowledge structures (e.g., graphs, retrieval contexts)
    \item Regime classification precedes generation
    \item System behavior is constrained by epistemic protocol, not model confidence
\end{itemize}

Validity-Aware AI enforces epistemic honesty structurally: systems must degrade their claims as entropy increases and must refuse predictive inference when regimes no longer support it.

Hallucination is reframed as a \textbf{protocol violation}, not a model failure.

\section{Core Unifying Principle}

Across all domains, the architecture enforces a single invariant:

\begin{quote}
\textit{No system may generate confident claims outside the regime where such claims are empirically valid.}
\end{quote}

This principle applies regardless of scale, substrate, or implementation.

\section{Scope and Implications}

The unified architecture does not seek to eliminate uncertainty or failure. Its goal is to make failure:

\begin{itemize}
    \item detectable
    \item bounded
    \item predictable
    \item and accountable
\end{itemize}

By integrating ontology, epistemology, regime dynamics, and implementation, the framework establishes \textbf{validity awareness} as a foundational requirement for responsible modeling and inference in complex systems.

\section*{References}

\noindent Magnusson, M. (2026). L0--L3 Regime Architecture in Entropy-Bounded Empiricism. Figshare. \url{https://doi.org/10.6084/m9.figshare.31112536}

\noindent Magnusson, M. (2025). Symbiosis: A Human--AI Co-Reflection Architecture Using Graph--Vector Memory for Long-Horizon Thinking. Figshare. \url{https://doi.org/10.6084/m9.figshare.30773684}

\noindent Magnusson, M. (2026). Entropy-Bounded Empiricism: SPARC175 Complete Documentation. GitHub/EFC.

\vfill
\noindent\rule{\textwidth}{0.4pt}
\small\textit{Unified Architecture Statement of: ``Validity-Aware AI: An Entropy-Bounded Architecture for Regime-Sensitive Inference'' (DOI: 10.6084/m9.figshare.31122970)}

\end{document}
